---
title: "Report of COSMIC data processing"
author: "Chuan-Xing Li"
date: '`r Sys.Date()`'
output: html_document
---

```{r, echo=FALSE, warning=FALSE,include=FALSE}
#setwd("/Users/cli/Documents/proj/copd/data")
## This is the main parameter input part.
#fpath <- "/Users/cli/Documents/proj/copd/data/" # fpath is the path of files. This is using the relative path. Before running this file, please set the path to the folder.
fpath <- "~/Documents/GitHub/ERJ_SNF/"
# fpath <- "/Users/cli/Documents/Projects/COPD/data/" # This is using the full path.
setwd(fpath)
fname <- paste(fpath,"rawdata/input_info.xlsx",sep="")
del_sample <- c(3208,3218) # The COSMIC ID will be deleted when inputing the data

## library required
library(knitr)
library(gdata)
library(ggplot2)
library(limma) # Quantile normalization of data (normalizeQuantiles )
library(corrplot)
source(paste0(fpath,"r/cosmic_process.R"))
source("./datasummary_format.R")
source('./iTRAQ_impute.R')
## input parameter sheet, maily for the name of data files and input paramteres.
colClass <- c(rep("character",9))
par <- read.xls(fname,sheet="par",colClasses = colClass,na.string = "NA")
rownames(par) <- par$item
class(par$nrows) <- "integer"
class(par$scol) <- "integer"
class(par$ecol) <- "integer"
rm(colClass)

tableNO <- 0; FigNO <- 0;
```

# Overview
This report is to recording the update of COSMIC data sets, and also some primary analysis results.   
**Version:** `r version$version.string`   
**Markdown file:** Report_COSMIC_Data.Rmd     
**Last update of Markdown file:** 2015-10-04

# Format of data summary and sample barcode
## Sample barcode
Format:  [Project]-[Participant]-[Gender]-[Smoking]-[Diagnosis]     
Example: COSMIC_1101_M_N_H

```{r, echo=FALSE, warning=FALSE}
tableNO <- tableNO + 1
t <- read.xls(fname,sheet="sample")
kable(t, caption =paste("Table ",tableNO,". Sample Barcode",sep=""),row.names = FALSE,escape = TRUE)
rm(t)
```

## Data platform Barcode
Format:  [Cell]-[Omic]-[Platform]    
Example: BAL_P_DIGE

```{r, echo=FALSE, warning=FALSE}
tableNO <- tableNO + 1
t <- read.xls(fname,sheet="platform")
kable(t, caption = paste("Table ",tableNO,". Data Platform Barcode",sep=""),row.names = FALSE)
rm(t)
```

```{r, echo=FALSE, include = FALSE, warning=FALSE}
# processing data or input processed data
if (file.exists("cosmic.RData")==FALSE) {
  
  cosmic<-cosmic_input(par,del_sample)
  fdate <- Sys.Date()
  cosmic$BAL_P_iTRAQ_ONE_impute<-iTRAQ_impute(cosmic$BAL_P_iTRAQ_ONE,10,"./BAL_P_iTRAQ_ONE") # imputation of iTRAQ data
  setwd(fpath)
  cosmic$BAL_P_iTRAQ_ALL_impute<-iTRAQ_impute(cosmic$BAL_P_iTRAQ_ONE,10,"./BAL_P_iTRAQ_ALL") # imputation of iTRAQ data
  setwd(fpath)
  save(cosmic,fdate,file="cosmic.RData")
  }
load("cosmic.RData")
```

# Information of raw dataset

```{r, echo=FALSE, warning=FALSE}
tableNO <- tableNO +1

temp <- fdate
title <- paste("Table ",tableNO, '. Update of COSMIC data files (',temp,')',sep='')
kable(par, caption = title,row.names = FALSE)
rm(title,temp,fname)
```

# Data Summary

All data are transformed to log2 based values.
```{r, echo=FALSE, warning=FALSE,fig.align='center'}
tableNO <- tableNO + 1
temp <- fdate
title <- paste("Table ",tableNO,' Summary of COSMIC data raw files (',temp,')',sep='')
omics <- c("HLA_typing","BAL_T_mRNA","BAL_P_DIGE","BAL_T_miR","BEC_T_miR","EXO_T_miR","BALF_M_Oxylip","Serum_M_Oxylip","Serum_M_Non_targeted","Serum_M_Biocrates","Serum_M_Kynurenine","Serum_M_Sphingolipid","BAL_P_iTRAQ_ONE_impute","BAL_P_iTRAQ_ALL_impute","BEC_P_TMT_impute")

# names(cosmic)[c(1,5,14,16,20)] # annotation files
t <- matrix(nrow = length(omics), ncol = 6)
colnames(t) <- c("item","nrow/features","ncol/sample","unique_features","missing_values","data type")
t <- as.data.frame(t)
t$item <-omics
sample <- rownames(cosmic$datasummary)
tsample <- matrix(nrow = length(sample),ncol = length(t$item)+2)
colnames(tsample) <- c(t$item, "cgroup","gender")
tsample <- as.data.frame(tsample)
tsample$cgroup <- cosmic$datasummary$cgroup
tsample$gender <- cosmic$datasummary$gender

for (i in 1:length(t$item)) {
  subject <- t$item[i]
  eval(parse(text=paste("data <- cosmic$",subject,sep="")))
  c <- match(colnames(data),cosmic$datasummary$barcode)
  data <- data[,is.na(c)==FALSE]
  t[i,2:3] <- dim(data)
  t[i,4] <- length(unique(rownames(data)))
  t[i,5] <- sum(sum(is.na(data)))
  t[i,6] <- class(data[1,1])
  tsample[,i] <- is.element(sample,colnames(data))+0
}
tsample$sum<- rowSums(tsample[,1:length(t$item)])
kable(t, caption = title,row.names = FALSE)
FigNO <- FigNO + 1

cols <- c(NH="#fc9272", SH="#a1d99b",SC="#fec44f",EC="#2162a6")
par(mar=c(5.1,4.1,8.1,10.1))
hist_cut <- ggplot(tsample, aes(x=sum, fill=cgroup),nclass=8)
hist_cut + geom_bar(binwidth=0.5)+scale_x_continuous(breaks=2.25:9.25,labels=2:9) + xlab("Number of Platforms") + ylab("Number of Samples")+scale_fill_manual(values = cols,name="Groups")+theme_bw() +ggtitle("A") +theme(plot.title = element_text(vjust = -50))
```
<center>**Figure `r FigNO`. The histogram for samples by platforms.**    
Red, green, yellow and blue class tags represent samples in groups of ***Non-smoking Healthy***, ***Smoking Healthy***, ***Smoking COPD*** and ***Ex-smoking COPD*** respectively. </center>

```{r, echo=FALSE, warning=FALSE,fig.align='center'}
FigNO <- FigNO + 1
# Heatmap
heat_raw<-tsample[,1:length(t$item)]
cgroup <- tsample$cgroup
rs <- rowSums(heat_raw)
cs <- colSums(heat_raw)
heat_raw <- heat_raw[order(-cs)] # reorder by the sum number of samples in each platform
h1 <- matrix(-1, nrow=nrow(heat_raw),ncol=1)
heat_raw <- cbind(as.numeric(cgroup)+1,as.numeric(tsample$gender)+5,h1,heat_raw)
heat <- heat_raw[with(heat_raw,order(cgroup,tsample$gender,-rs)),]
heat <- t(heat)
rownames(heat)[1]<-paste("Groups (",ncol(heat),")",sep="")
rownames(heat)[2]<-"Gender"
rownames(heat)[3]<-""
for (i in 4:nrow(heat)){
        rownames(heat)[i] <- paste(rownames(heat)[i]," (",sum(heat[i,]),")",sep="")
}


palette.breaks <- seq(-1,8,by=1) - 0.5
color.palette  <- colorRampPalette(c("white", "#deebf7","#67a9cf","#fc9272","#a1d99b","#fec44f","#2162a6","#344152","grey"))(length(palette.breaks) - 1)

heatmap(heat[nrow(heat):1,], Rowv = NA, Colv = NA, scale = "none", labCol =NA, distfun = "binary", col = color.palette, breaks = palette.breaks)
```
<center>**Figure `r FigNO`. Summary of Multi-omics Data from COSMIC Project.**   
Each row represents a platform, and each column is a sample. Dark and light blue cells shows the data is available or not. Red, green, yellow and blue class tags represent samples in groups of ***Non-smoking Healthy***, ***Smoking Healthy***, ***Smoking COPD*** and ***Ex-smoking COPD*** respectively. Black and grey in ***Gender*** bar show female and male respectively. MF: alveolar macrophages; BEC: bronchial epithelial cell; BAL: bronchoalveolar lavage; EXO: exosomes from bronchoalveolar lavage fluid (BALF). The number in brackets is how many samples are tested in each platform. The barcode of platform see "Data platform Barcode" part.</center>
&nbsp;

```{r, echo=FALSE, warning=FALSE,fig.align='center'}     
# heatmap of correlation between platforms
FigNO <- FigNO +1
heat_raw <- tsample[,t$item]
hamming.distance <- function(x){
        cn <- colnames(x)
        x <- as.matrix(x)
        y <- matrix(0,ncol(x),ncol(x))
        for (i in 1:(ncol(x)-1)) {
                y[i,i] = sum(x[,i])
                for (j in (i+1):ncol(x)) {
                        t <- cbind(x[,i],x[,j])
                        y[i,j] <- sum(rowSums(t)==2)
                        y[j,i] <- y[i,j]
                }
        }
        rownames(y) <- cn
        colnames(y) <- cn
        y[ncol(y),ncol(y)] <- sum(x[,ncol(x)])
        y/nrow(x)
}
corr.m<-hamming.distance(heat_raw)

corrplot(corr.m,order="AOE",p.mat = corr.m*nrow(heat_raw), cl.lim=c(0,1),sig.level=0.01,insig="p-value",tl.col="black")
```

<center> **Figure `r FigNO`. Correlation map between different platforms**   
The concentration of color is positive correlated with the percent of samples tested by both platforms. The values are number of samples tested by the corresponding platform or both platforms. </center>

# Boxplot of all platform
```{r, echo=FALSE, warning=FALSE,fig.align='center'}  
subjectT <- c("BAL_T_mRNA","BAL_T_miR","BEC_T_miR","EXO_T_miR","Serum_M_Non_targeted")
subjectT2 <- c("BALF_M_Oxylip","Serum_M_Oxylip","Serum_M_Biocrates","Serum_M_Kynurenine","Serum_M_Sphingolipid","BAL_P_DIGE","BAL_P_iTRAQ_ONE_impute","BAL_P_iTRAQ_ALL_impute","BEC_P_TMT_impute")

        for (i in 1:length(subjectT)) {                
                subject <- subjectT[i]

                eval(parse(text = paste("data = cosmic$",subject,sep="")))
                c <- match(colnames(data),cosmic$datasummary$barcode)
                data <- data[,is.na(c)==FALSE]
                    boxplot(data,xaxt="n",xlab="Samples",ylab="log2(values)",main=subject)    
        }
for (i in 1:length(subjectT2)) {                
                subject <- subjectT2[i]

                eval(parse(text = paste("data = cosmic$",subject,sep="")))
                c <- match(colnames(data),cosmic$datasummary$barcode)
                data <- data[,is.na(c)==FALSE]
                   boxplot(data,xaxt="n",xlab="Samples",ylab="Raw values/Ratio",main=subject)   
        }

```

## Data Summary
### Before importing to R
1. Copy sheet ***Data block Incl Excl_QC excl*** to sheet ***data***
2. Delet contents and finally as in sheet ***data***
### After importing to R
3. Using *datasummary_format.R* to formated barcode and 4 groups in ***cgroup*** column as NH, SH, SC and EC. 

## Clinical Data 2016-03-31
### Before importing to R
1. Add sheet ***data*** and copy content from ***COSMIC clinical 2016-03-31*** to it;
2. Replace empty cell as ***NA*** (n=388); replace ***na*** as ***NA***(n=541)
3. Duplicate the first row ***Subject ID*** to ***COSMIC ID*** at row 2;
4. Copy the first column to the front, replace ***space*** to ***_***, rename ***Smoking_status*** to ***Smoking*** and ***Diagnosis_(Healthy=1,__COPD=2)*** to ***Diagnosis***.

## clinical data for bioconductor prepared by Vincenzo (cosmic$clinic_bioconductor)
The code is in preprocessing.R. After input COSMIC clinical_2016-03-31_selected.csv, a column variable named barcode is added.

### After importing to R
5. Name column name as datasummary$barcode
6. Use the first column as rowname, then delet the first column

## Clinical Data 
### Before importing to R
1. Add sheet ***data*** and copy content from ***clinical_2015...*** to it;
2. Replace empty cell as ***NA*** (n=248); replace ***na*** as ***NA***(n=684); Delete column after ***DT***
3. Replace ***patient no*** to ***COSMIC ID*** at A184;
4. Copy the first column to the front, replace ***space*** to ***_***, rename ***Smoking_*** to ***Smoking*** and ***Doctor_*** to ***Doctor***.

### After importing to R
5. Rename duplicated row names to "rawname" + 2
6. Name column name as datasummary$barcode
7. Use the first column as rowname, then delet the first column

## BAL_T_mRNA
### Before importing to R
1. Copy sheet "mRNA_all subj o genes_quant.xls" to "data"
2. Delete row 2 to 16; Delet column B, G to K; Delet column without expression values (empty cells)
3. Insert a copy of first row as the second row. The first item was renamed as "COSMIC_ID"; formatted the COSMIC ID.
4. No missing values from column E to BD
5. Number of row = 41002, start column = 6
6. Save sheet "data" to .csv file ***`r par["BAL_T_mRNA","fname"]`***

## BAL_P_DIGE
### Before importing to R
1. Copy pure data to new sheet "data", column E, Q to CO
2. Insert a copy of first row as the second row. The first item was renamed as "COSMIC_ID"

### After importing to R
3. Convert the value from the original **log10** tranformation to **log2** transformation

## BAL_P_iTRAQ
### Before importing to R
1. copy sheet ***iTRAQ_BALc_75p ALL_ratio_log2*** to ***dataALL***
2. Delete row 2 to 4; delete column B to S; for row 1 from column B to end replace "_log2*" to empty and formated cell as number.
3. replace all empty cell to "NA". n=1599
4. copy sheet ***iTRAQ_BALc_75per_ONE_ratio_log2*** to ***dataONE***
5. Delete row 2 to 4; delete column B to S; for row 1 from column B to end formated cell as number.
6. replace all empty cell to "NA". n=20652

### After importing to R
7. BAL_P_iTRAQ_ALL and BAL_P_iTRAQ_ONE correspond to step 1 and 3 respectively.
8. Test different KNN imputation parameters based on BAL_P_iTRAQ_ONE
9. The test from KNN imputation is not good, still impute with K = 10. The report is in "./doc/Report_Imputation_BAL_iTRAQ_CXL160525.docx" using iTRAQ_impute.R

## BAL_T_miR
### Before importing to R
1. Copy sheet "HsamiR_quantle_BAL_excl QCoutl" to "data"
2. Insert a copy of first row as the second row. 
3. Fill the "input_info.xlsx" par["BAL_T_miR","scol"]<-43, starting column of expression data.
4. Change the cell of H2 (row 2, column aveA) to 1.00
### After importing to R
5. Name rownames as column "Probe.Sequence".
6. Delete column 1 to par["BAL_T_miR","scol"]<-43 -1, except column *aveA* which is moved to the last column.
7. Formated columnames as formated COSMIC ID, delete the columns without formated COSMIC ID.

## BEC_T_miR
### Before importing to R
1. Copy sheet "Hsa_miR_Quantile" to "data".
2. Delete all column not sample expression except Prob.sequence and aveA

### After importing to R
5. Name rownames as column "Probe.Sequence".
6. Delete column 1 to par["BEC_T_miR","scol"]<-43 -1, except column *aveA* which is moved to the last column.
7. Formated columnames as formated COSMIC ID, delete the columns without formated COSMIC ID.

## EXO_T_miR
### Before importing to R
1. Copy sheet "EXO_miR_quantileAll_log2_66 sub" to "data".
2. Insert a copy of first row as the second row. 
3. Fill the "input_info.xlsx" par["EXO_T_miR","scol"]<-43, starting column of expression data.
4. Change the cell of H2 (row 2, column aveA) to 1.00.
### After importing to R
5. Name rownames as column "Probe.Sequence".
6. Delete column 1 to par["EXO_T_miR","scol"]<-43 -1, except column *aveA* which is moved to the last column.
7. Formated columnames as formated COSMIC ID, delete the columns without formated COSMIC ID.

## BALF_M_Oxylip

### Before importing to R
1. Replace empty cell to "NA",n=226
### After importing to R
2. Using "SecID" as rownames
3. Delete column/sample with all NA values
4. Imputation of missing values as 1/3 of LLOQ of special features

## Serum_M_Oxylip
### Before importing to R
1. Replace empty cell to "NA",n=222
### After importing to R
2. Using "SecID" as rownames
3. Delete column/sample with all NA values
4. Imputation of missing values as 1/3 of LLOQ of special features

## Oxylip_annot
### Before importing to R
1. Fill A1=SecID, B1=Symbol
2. Replace empty cell, "-" to "NA"
### After importing to R
3. First row as head/columnames
4. Using "SecID" as rownames

## Serum_M_Non_targeted
### Before importing to R
1. Copy sheet "Non-targeted(Q-Ex)" to data with transpose
2. Correct COS_1101_2206b to COS_1101; COS_2201_2210b to COS_2201
3. Delete "COS_"" in row 1

### After importing to R
4. Convert the raw non-log transformed value to **log2** transformation

## Non_target annot
### Before importing to R
1. The same as Serum_M_Non_targeted, from Serum_metabolomics_3 platform_Shama_2016-05.xlsx sheet "data"/"Non-targeted(Q-Ex)"

### After importing to R
2. The first 8 columns are extracted
3. Merge **conversionTable** by Vincenzo into it, by adding a new column "KEGG"



## Serum_M_Biocrates, Serum_M_Kynurenine, Serum_M_Sphingolipid
### Before importing to R
1. Replace "<LLOQ" and empty cell as "NA"

### After importing to R
2. Transpose data into row as features and column as samples
3. Filter features with only NA or 0

### BEC_P_TMT
### Before importing to R
1. Copy sheet **TMT_BEC_75%_All_log2_ratio** to **data**
2. In sheet **data**, delete column B to S, delete row 1, 3 and 4
3. Replace empty cell as "NA", n = 2234

### After importing to R
4. Transfer first row to COSMIC barcode as column name; use first column as row name
